---
title: "Examining the Correlation between Neighbourhood Property Crime Rates and Social Factors in Toronto, ON"
#subtitle: "A subtitle"
#thanks: "Paper submitted as the final project of ENVSOCTY 4GA3: Applied Spatial Statistics."
author:
- name: Ashmitha Bala
  student_number: 400148743
- name: Cassis Lindsay
  student_number: 400194383
- name: CJ Murphy
  student_number: 400195804
- name: Hayat Shariff
  student_number: 400258691
- name: Zoe Spronk
  student_number: 400231466
subject: "ENVSOCTY 4GA3"
abstract: "This paper reports our analysis of property crime rates in Toronto, ON and their correlation with population density, age, race, and income by neighbourhood. Data were retrieved from The City of Toronto Open Data Portal."
keywords: "crime, social factors, spatial analysis"
date: "4/22/2020"
output:
  pdf_document:
    # The project-template-default.tex file was minimally adapted from Steven V. Miller's template for academic manuscripts. See:
    # http://svmiller.com/blog/2016/02/svm-r-markdown-manuscript/
    # https://github.com/svmiller/svm-r-markdown-templates/blob/master/svm-latex-ms.tex
    template: project-template-default.tex
bibliography: [bibliography.bib, packages.bib]
always_allow_html: true
---

```{r clear, include=FALSE}
#clearing the environment 
rm(list = ls())
```

<!--Chunks of code can have names; the chunk option "include" controls whether the chunk and its output are printed in the final document-->
```{r load-packages, include=FALSE}
# only run the following line if the package is not already installed
# install.packages("opendatatoronto")
library(dplyr) # for accessing Toronto Open Data
library(opendatatoronto) # for accessing Toronto Open Data
library(gridExtra) # for faceting with ggplot
library(plotly) # interactive plotting
library(sf) # working with simple features data
library(spdep) # spatial regression analysis
library(tidyverse) # data manipulation, analysis, and visualization (incl. ggplot2, magrittr (pipe operators), etc)
library(viridis) # For colour palettes (colour blind friendly)
```


```{r write-package-bib, include=FALSE}
# This function is used to write a bibliography for the `R` packages used in the paper
knitr::write_bib(file = 'packages.bib')
```


```{r get-crime-data, include = FALSE}
# get package
package <- show_package("fc4d95a6-591f-411f-af17-327e6c5d03c7")
package
 
# get all resources for this package
resources <- list_package_resources("fc4d95a6-591f-411f-af17-327e6c5d03c7")
 
# identify datastore resources; by default, Toronto Open Data sets datastore resource format to CSV for non-geospatial and GeoJSON for geospatial resources
datastore_resources <- filter(resources, tolower(format) %in% c('csv', 'geojson'))
 
# load the first datastore resource as a sample
TO_crime <- filter(datastore_resources, row_number()==1) %>% get_resource()
```

```{r aggregating-property-crime-rates, include = FALSE}
# creating new columns in the df with aggregated property crime rates by year
# auto theft + breaking and entering + robbery + theft over 5k
# Rates are per 100k
TO_crime <- TO_crime %>%
  mutate(propertyCrime2014 = ((AutoTheft_2014 + BreakAndEnter_2014 + Robbery_2014 + TheftOver_2014)/F2020_Population_Projection)*100000,
         propertyCrime2015 = ((AutoTheft_2015 + BreakAndEnter_2015 + Robbery_2015 + TheftOver_2015)/F2020_Population_Projection)*100000,
         propertyCrime2016 = ((AutoTheft_2016 + BreakAndEnter_2016 + Robbery_2016 + TheftOver_2016)/F2020_Population_Projection)*100000,
         propertyCrime2017 = ((AutoTheft_2017 + BreakAndEnter_2017 + Robbery_2017 + TheftOver_2017)/F2020_Population_Projection)*100000,
         propertyCrime2018 = ((AutoTheft_2018 + BreakAndEnter_2018 + Robbery_2018 + TheftOver_2018)/F2020_Population_Projection)*100000,
         propertyCrime2019 = ((AutoTheft_2019 + BreakAndEnter_2019 + Robbery_2019 + TheftOver_2019)/F2020_Population_Projection)*100000,
         propertyCrime2020 = ((AutoTheft_2020 + BreakAndEnter_2020 + Robbery_2020 + TheftOver_2020)/F2020_Population_Projection)*100000)
         
#converting all character data to numeric
TO_crime$propertyCrime2014<- as.numeric(TO_crime$propertyCrime2014)
TO_crime$propertyCrime2015<- as.numeric(TO_crime$propertyCrime2015)
TO_crime$propertyCrime2016<- as.numeric(TO_crime$propertyCrime2016)
TO_crime$propertyCrime2017<- as.numeric(TO_crime$propertyCrime2017)
TO_crime$propertyCrime2018<- as.numeric(TO_crime$propertyCrime2018)
TO_crime$propertyCrime2019<- as.numeric(TO_crime$propertyCrime2019)
TO_crime$propertyCrime2020<- as.numeric(TO_crime$propertyCrime2020)
#finding yearly means of all neighbourhood property crime rates
meanprop2014 <- mean(TO_crime$propertyCrime2014)
meanprop2015 <- mean(TO_crime$propertyCrime2015)
meanprop2016 <- mean(TO_crime$propertyCrime2016)
meanprop2017 <- mean(TO_crime$propertyCrime2017)
meanprop2018 <- mean(TO_crime$propertyCrime2018)
meanprop2019 <- mean(TO_crime$propertyCrime2019)
meanprop2020 <- mean(TO_crime$propertyCrime2020)
#joining yearly means of property crime rates
TO_prop_means <- c(meanprop2014,meanprop2015,meanprop2016,meanprop2017,meanprop2018,meanprop2019,meanprop2020)
```

```{r get-demographic-data, include = FALSE}
	
# loading social package
package <- show_package("6e19a90f-971c-46b3-852c-0c48c436d1fc")
package
 
# get all resources for this package
resources <- list_package_resources("6e19a90f-971c-46b3-852c-0c48c436d1fc")
 
# identify datastore resources; by default, Toronto Open Data sets datastore resource format to CSV for non-geospatial and GeoJSON for geospatial resources
datastore_resources <- filter(resources, tolower(format) %in% c('csv', 'geojson'))
 
# load the first datastore resource as a sample
TO_social <- filter(datastore_resources, row_number()==1) %>% get_resource()
```

```{r reformatting-demographic-data, include = FALSE}
#converting rows as columns and columns as rows
TO_social <- data.frame(t(TO_social[-1]))
#assigning column names from data_1
colnames(TO_social) <- TO_social[4,]
#deleting first rows in TO_social
TO_social <- TO_social[-c(1:4),]
```

```{r joining-demo-data-to-crime-df, include = FALSE}
# Adding demographic data of interest to the crime df (which has geometry) so that it is mappable
# Neighbourhood names = shared between the 2 datasets
# First, we need to arrange TO_crime to be alphabetical by neighbourhood name so that we can match the 2 datasets
# removing apostrophe from "L'amoreaux" so that it follows the same alphabetical convention as TO_social
TO_crime[66,3] <- "Lamoreaux"
# arranging TO_crime alphabetically by neighbourhood
TO_crime <- TO_crime %>% arrange(Neighbourhood)
# Next, extract the columns of interest from TO_social into vectors
# Problem: pop density, income and race are not currently numeric, need to remove commas and convert
TO_social[2:141,8] <- as.numeric(gsub(",","",TO_social[2:141,8])) #pop density
TO_social[2:141,1129] <- as.numeric(gsub(",","",TO_social[2:141,1129])) #income
TO_social[2:141,1340] <- as.numeric(gsub(",","",TO_social[2:141,1340])) #race
TO_social[2:141,3] <- as.numeric(gsub(",","",TO_social[2:141,3])) #2016 population
population_density <- as.numeric(TO_social[2:141, 8]) # pop density per square km
total_low_income <- as.numeric(TO_social[2:141, 1129]) # number of identified low income individuals after tax
total_visible_minority <- as.numeric(TO_social[2:141, 1340]) # number of inhabitants who belong to visible minority groups
population_2016 <- as.numeric(TO_social[2:141, 3]) # population in 2016
# The age data requires a bit more work
# we are going to sum all the columns, male and female, between the ages of 15-29 into one single vector, then use that for analysis in the crime df
Young <- as.numeric(TO_social[2:141,20]) + as.numeric(TO_social[2:141,21]) + as.numeric(TO_social[2:141,22]) + as.numeric(TO_social[2:141,41]) + as.numeric(TO_social[2:141,42]) + as.numeric(TO_social[2:141,43])
# Finally, join these vectors into the TO_crime df
TO_crime$population_density <- population_density
TO_crime$total_low_income <- total_low_income
TO_crime$total_visible_minority <- total_visible_minority
TO_crime$total_pop_aged_15to29 <- Young
TO_crime$population_2016 <- population_2016
```

<!--This create a page break, i.e., starts a new page-->
<!--\newpage-->

<!-- 
To cite references in your bibliography.bib file, use [@item] if you want it to be cited in brackets, or @item if you want it to be cited as Name (year). If you want to cite various items in brackets, separate them with semicolons [@item1; @item2]
-->

<!--Use "#" for section headers-->
<!-- Use ** on both sides for bold text, and * on both sides for italics -->

# Introduction

Insert Introduction Here

# Background

Insert Background Here

# Study area

The relationship between property crime rates and social factors were studied across the census neighbourhoods of Toronto, ON, Canada (see Figure \ref{fig:toronto-neighbourhoods}). There are 140 neigbourhoods in total at the time of this study. *write more about why this area was selected*.

```{r toronto-neighbourhoods, echo = FALSE, fig.cap={"\\label{fig:toronto-neighbourhoods} Census Neighbourhoods of Toronto, ON, Canada"}}
# Map of Toronto neighbourhoods
# Colour extracted from Viridis default palette to remain consistent with other figures produced
ggplot(TO_crime) + 
  geom_sf(color = "black", fill = "#5DC863FF", size = .3) + 
  theme_minimal()
```

# Data

The data used in this study comes from the City of Toronto Open Data Portal, accessed using the **opendataportal** package in R. [@R-opendatatoronto]. Two separate datasets were used: *Neighbourhood Crime Rates* and *Neighbourhood Profiles*. The *Neighbourhood Crime Rates* dataset reports both the raw count and rates per 100,000 of seven different types of crime within the 140 neighbourhoods of Toronto. [@TorontoPoliceData]. Of the seven types of crime, the four reported types of property crime — auto theft, breaking and entering, robbery, and theft over $5000 — were the focus of this study. The *Neighbourhood Profiles* dataset reports the demographic, social, and economic findings of the 2016 census for each of Toronto's 140 neighbourhoods. [@statsCanCensus2016]. The data regarding population density, age, income, and race were utilised as the social factors of interest for this study.

# Methods

Insert methods here

# Results

<!--background info maps about geographic demographics in Toronto -->


```{r pop-density-map, echo = FALSE, fig.cap={"\\label{fig:pop-density} Population density by neighbourhood (2016)"}}
# Map of population density
popDensity <- ggplot() +
  geom_sf(data = TO_crime,
          aes(fill = population_density),
          colour = "white") +
  scale_fill_viridis(direction = -1,
                     option = "D",
                     name = "People per square km") +
  theme_void()
popDensity
```

```{r low-income-percent-map, echo = FALSE, fig.cap={"\\label{fig:low-income-percent} Percentage of inhabitants who have been identified as 'Low Income After Taxation' by neighbourhood (2016)"}}
# add new column w/percent of individuals who are low income
TO_crime <- TO_crime %>%
  mutate(low_income_percent = (100/population_2016)*total_low_income)
# Map of low income percent
LowIncomePercent <- ggplot() +
  geom_sf(data = TO_crime,
          aes(fill = low_income_percent),
          colour = "white") +
  scale_fill_viridis(direction = -1,
                     option = "D",
                     name = "Percent") +
  theme_void()
LowIncomePercent
```

```{r visible-minority-percent-map, echo = FALSE, fig.cap={"\\label{fig:visible-minority-percent} Percentage of inhabitants who belong to visible minority groups by neighbourhood (2016)"}}
# add new column w/percent of individuals who are visible minorities
TO_crime <- TO_crime %>%
  mutate(visible_minority_percent = (100/population_2016)*total_visible_minority)
# Map of visible minorty percent
visibleMinorityPercent <- ggplot() +
  geom_sf(data = TO_crime,
          aes(fill = visible_minority_percent),
          colour = "white") +
  scale_fill_viridis(direction = -1,
                     option = "D",
                     name = "Percent") +
  theme_void()
visibleMinorityPercent
```

```{r young-percent-map, echo = FALSE, fig.cap={"\\label{fig:young-percent} Percentage of inhabitants aged 15-29 by neighbourhood (2016)"}}
# add new column w/percent of individuals who are aged 15 to 29
TO_crime <- TO_crime %>%
  mutate(percent_aged_15to29 = (100/population_2016)*total_pop_aged_15to29)
# Map of percent aged 15 to 29
youngPercent <- ggplot() +
  geom_sf(data = TO_crime,
          aes(fill = percent_aged_15to29),
          colour = "white") +
  scale_fill_viridis(direction = -1,
                     option = "D",
                     name = "Percent") +
  theme_void()
youngPercent
```

<!--Maps of property crime rates -->

```{r 2020-split-crime, echo = FALSE, fig.cap={"\\label{fig:split-property-crime-rate-2020} Four categories of property crime rates by neighbourhood (2016)"}}
# Maps of the 4 components of total property crime in 2016
# Auto theft rate map
auto2016 <- ggplot() +
  geom_sf(data = TO_crime,
          aes(fill = AutoTheft_Rate2016),
          colour = "white") +
  scale_fill_viridis(direction = -1,
                     option = "D",
                     name = "Rate per 100k") +
  theme_void() +
  ggtitle("Auto Theft") +
  theme(plot.title = element_text(hjust = 0.5)) #centers the title
# breaking and entering map
break_enter2016 <- ggplot() +
  geom_sf(data = TO_crime,
          aes(fill = BreakAndEnter_Rate2016),
          colour = "white") +
  scale_fill_viridis(direction = -1,
                     option = "D",
                     name = "Rate per 100k") +
  theme_void() +
  ggtitle("Breaking and Entering") +
  theme(plot.title = element_text(hjust = 0.5)) #centers the title
# robbery map
robbery2016 <- ggplot() +
  geom_sf(data = TO_crime,
          aes(fill = Robbery_Rate2016),
          colour = "white") +
  scale_fill_viridis(direction = -1,
                     option = "D",
                     name = "Rate per 100k") +
  theme_void() +
  ggtitle("Robbery") +
  theme(plot.title = element_text(hjust = 0.5)) #centers the title
# theft over $5000 map
theftOver2016 <- ggplot() +
  geom_sf(data = TO_crime,
          aes(fill = TheftOver_Rate2016),
          colour = "white") +
  scale_fill_viridis(direction = -1,
                     option = "D",
                     name = "Rate per 100k") +
  theme_void() +
  ggtitle("Theft Over $5000") +
  theme(plot.title = element_text(hjust = 0.5)) #centers the title
# presentation of maps in one figure
grid.arrange(auto2016, break_enter2016, robbery2016, theftOver2016)
```

<!--all factors are combined below into aggregated property crime rates -->

```{r 2016-property-crime, echo = FALSE, fig.cap={"\\label{fig:property-crime-rate-2016} Aggregated property crime rates by neighbourhood (2016)"}}
# Map of aggregated property crime rates of 2016
# (auto theft, breaking and entering, robbery, and theft over 5k)
propCrime2016 <- ggplot() +
  geom_sf(data = TO_crime,
          aes(fill = propertyCrime2016),
          colour = "white") +
  scale_fill_viridis(direction = -1,
                     option = "D",
                     name = "Rate per 100k") +
  theme_void()
propCrime2016
```

<!--analytical maps and figures -->

```{r crime-city-mean, echo = FALSE, fig.cap={"\\label{fig:crime-city-mean} Neighbourhood crime rates relative to mean crime rate of the whole of Toronto (2016)"}}
# mapping neighbourhoods with high and low crime rates (above or below the city neighbourhood mean)
# meanprop2016 # use to check value for ifelse
TO_crime$propertyCrime2016TF <- ifelse(TO_crime$propertyCrime2016 > 548.8508, "TRUE", "FALSE")
propCrime2016TF <- ggplot() + 
  geom_sf(data = TO_crime, 
          aes(fill = propertyCrime2016TF),
          colour = "white") + 
  labs(fill = "Crime Rate Above City Mean") +
  theme_void()
propCrime2016TF
```

```{r spatial-weight-matrix, include = FALSE}
#creating spatial weight matrix for TO_crime data
TO_crime.nb <- poly2nb(pl = TO_crime)
TO_crime.w <- nb2listw(TO_crime.nb)
```

```{r moran-test-crime, include = FALSE}
#executing Moran's test to test crime rates for spatial autocorrelation
moran.test(TO_crime$propertyCrime2016, TO_crime.w)
# p value = 8.456e-10, therefore not random 
```
```{r moran-plot-crime, echo = FALSE, fig.cap={"\\label{fig:moran-plot-crime} Moran's plot for spatial autocorrelation regarding neighbourhood property crime rates (2016)"}}
moran.plot(TO_crime$propertyCrime2016, TO_crime.w, xlab = "Property Crime Rate 2016 (per 100k)", ylab = "Spatially Lagged Property Crime Rate 2016") 
#clustered in the bottom left and top right = spatial autocorrelation. High and low crime rates are similar to their neighbours. Therefore, crime has spatial autocorrelation in Toronto. **write this better**
```

```{r creation-and-summary-of-regression-models, include = FALSE}
# regression models of property crime rates and various social factors. All data from 2016.
# income
model1 <- lm(formula = TO_crime$propertyCrime2016 ~ TO_crime$low_income_percent, data = TO_crime)
summary(model1)
# race
model2 <- lm(formula = TO_crime$propertyCrime2016 ~ TO_crime$total_visible_minority, data = TO_crime)
summary(model2)
# age
model3 <- lm(formula = TO_crime$propertyCrime2016 ~ TO_crime$percent_aged_15to29, data = TO_crime)
summary(model3)
# population density
model4 <- lm(formula = TO_crime$propertyCrime2016 ~ TO_crime$population_density, data = TO_crime)
summary(model4)
```

```{r crime-vs-income, echo = FALSE, results='hide',fig.keep='all', fig.cap={"\\label{fig:crime-income} Regression model comparing percentage of low income individuals and property crime rates by neighbourhood (2016)"}}
plot(TO_crime$propertyCrime2016 ~ TO_crime$low_income_percent, xlab = "Percentage of Low Income Individuals", ylab = "Neighbourhood Property Crime Rate", pch = 16) + abline(model1, col = "red")
```

```{r crime-vs-race, echo = FALSE, results='hide',fig.keep='all', fig.cap={"\\label{fig:crime-race} Regression model comparing percentage of visible minority individuals and property crime rates by neighbourhood (2016)"}}
plot(TO_crime$propertyCrime2016 ~ TO_crime$visible_minority_percent, xlab = "Percentage of Visible Minority Individuals", ylab = "Neighbourhood Property Crime Rate", pch = 16) + abline(model2, col = "red")
```

```{r crime-vs-age, echo = FALSE, results='hide',fig.keep='all', fig.cap={"\\label{fig:crime-age} Regression model comparing percentage of young (aged 15-29) individuals and property crime rates by neighbourhood (2016)"}}
plot(TO_crime$propertyCrime2016 ~ TO_crime$total_pop_aged_15to29, xlab = "Percentage of Population Aged 15-29", ylab = "Neighbourhood Property Crime Rate", pch = 16) + abline(model3, col = "red")
```

```{r crime-vs-density, echo = FALSE, results='hide',fig.keep='all', fig.cap={"\\label{fig:crime-density} Regression model comparing population density and property crime rates by neighbourhood (2016)"}}
plot(TO_crime$propertyCrime2016 ~ TO_crime$population_density, xlab = "Neighbourhood Population Density", ylab = "Neighbourhood Property Crime Rate", pch = 16) + abline(model4, col = "red")
```


# Analysis

Insert Analysis Here

# Conclusion

Insert Conclusion Here

# References
